Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-02-11 doc/article.tex
Greg's comments in

@@

@@

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted peculiar velocity power spectrum $k^3P_{vv}(z=0)$ for $\mu=1, 0.5$ (solid black, cyan)  as predicted for General Relativity in the linear regime.
Overplotted are volume-weighted peculiar velocity shot-noise  $k^3\sigma^2/n$ at $z=0.1$ expected from 2- and 10-year (red, black) supernova densities, 0.08 and 0.15~mag (dashed, dotted) intrinsic magnitude dispersions.
The bottom solid horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{max}$.
\label{power:fig}}
\end{figure}

@@

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields  absolute magnitude
dispersion $\sigma_M \gtrsim 0.12$~mag.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}. 
Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.  
At this precision the intrinsic velocity dispersion  at $z=0.028$ is  of $300$~km\,s$^{-1}$, i.e.\ a single SN~Ia  is of such quality as to
provide $S/N \sim 1$.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.  With similar densities,
 the 10-year SN~Ia survey will have
 a 29$\times$ reduction in shot-noise ($\sigma^2_M/n$) relative to the Fundamental Plane survey of TAIPAN.

@@

We project uncertainties  on the growth index, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}
\citep[there is an alternative approach using an estimator for the mean pairwise velocity][]{2011PhRvD..83d3004B}.
***Super-sample vairaince***
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_{r_{min}}^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the parameters considered are $\lambda \in \{\gamma,bD, \Omega_{M0}\}$.
The parameter dependence enters through $(fD)(\gamma)$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
Both $f$ and $D$ depend on $\Omega_M=\frac{\Omega_{M0}}{\Omega_{M0} + (1-\Omega_{M0})a^3}$.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered.

@@

We consider peculiar velocity surveys for a range of redshift depths $z_{max}$ for survey durations of $t=2$ and 10~years.
The other survey parameters
$\Omega=3\pi$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
The $k$-limits are  taken to be $k_{min} = \pi/r_{max}$  and 
$k_{max} = 0.1$~$h$\,Mpc$^{-1}$.
A minimum distance $r_{min}=r(z=0.01)$ is imposed as the physics derivation is taken at the limit
where peculiar velocities are significantly  smaller than
the cosmological redshift.
The sample-selection efficiency $\epsilon$ is redshift-independent, i.e.\ the native redshift distribution is not sculpted.
The input bias of SN~Ia host galaxies is set as $b=1.2$.  An independent measurement of $\Omega_{M0}=0.3\pm0.005$
is included and is important contributor to the $\gamma$ constraint.  ***Uniform sampling ***

@@

{\bf Host-galaxies Need Precise Redshifts:} Peculiar velocities come from the difference between host-galaxy and cosmological redshifts.  Host redshift uncertainties need to be
 $<0.5$\% in order not to contribute significantly
to the error budget.  We thus conclude that there is the need for spectroscopic $R>200$ host-galaxy redshifts.  These redshifts could come naturally
from galaxy signal in  the spectroscopic transient classification if the spectrograph had sufficient resolution.  ***Check number and explain. ***  Alternatively, 
moderate-resolution (bright-)galaxy  redshift surveys (e.g.\ DESI, 4MOST) can  provide multiplexed observations of 
a significant fraction of nearby SN hosts before  discovery or after transient light has faded away.

@@

@@

%SNe~Ia are already powerful probes of the homogeneous cosmological expansion of the Universe.  
In the next decade,
the high number of SN discoveries together with improved precision in their distance precisions will make $z<0.2$ SNe~Ia, more so than
galaxies,  powerful probes of gravity through their effect  on the growth of structure. 
A peculiar velocity survey must discover, classify, and get distances of SNe~Ia, as well as obtain redshifts of their host galaxies.
Choices in survey design impact the precision obtained on the growth index $\gamma$.  
No other probe of growth of structure or tracer of peculiar velocity can alone provide comparable precision on  $\gamma$ in the next decade.
Combined low-redshift peculiar velocity and high-redshift RSD $fD$ measurements are highly complimentary as together they probe the
$\gamma$-dependent shape of $fD(z)$ (not just its normalization) and potential scale-dependent influence of gravitational models, since low-
and high-redshift surveys are weighted by lower and higher $k$-modes respectively.
SN~Ia peculiar velocity surveys are of the highest scientific
interest and we encourage
the community to develop aggressive surveys in the pursuit of testing  General Relativity and probing gravity.

***
DESI and 4MOST will exist in the 2020 decade, (the former having been endorsed by Astro 2010), but at low redshift their constraints on fs8 will be limited by cosmic variance, Since these two redshifts surveys will cover most of the usable sky, they will saturate the RSD technique, making PVs the only way to improve on fs8 at low z. TAIPAN and DESI BGS will be able to measure FP distances for nearly all usable galaxies nearby, so at low z the FP PV technique will also saturate in the coming decade. Thus, SN PV will be the most powerful way to further constrain $\sigma_\gamma$. Moreover, SN PV directly benefits from DESI and 4MOST through the cross-correlation of the galaxy density and velocity fields, and the in the redshifts provided for many of the SN host galaxies.

t TAIPAN+(+4MOST)+DESI are *capable* of saturating the FP technique nearby. If 4MOST or DESI chose not to do FP or fail in the attempt, that is just more reason to do SN PVs! But I think it's a good idea for readers to know that SN PVs are needed even if 4MOST and DESI do the best job they can on FP, simply because one runs out of suitable nearby FP galaxies.
***
%\begin{comment}
\newpage
We are interested in
$$F^{-1}_{00,\alpha} = \frac{F_{11,\alpha}}{F_{00}F_{11}-F_{01}^2} - \frac{F_{11}}{(F_{00}F_{11}-F_{01}^2)^2}\left(F_{00,\alpha}F_{11}+
F_{00}F_{11,\alpha} - 2 F_{01}F_{01,\alpha}\right).$$

>>>>

\author{A.~Bahmanyar}
\affiliation{Department of Astronomy and Astrophysics, University of Toronto,
50 St. George Street, Toronto, Ontario, Canada M5S 3H4}

@@

\author{R.~Hlo\u{z}ek}
\affiliation{Department of Astronomy and Astrophysics, University of Toronto,
50 St. George Street, Toronto, Ontario, Canada M5S 3H4}

@@

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted peculiar velocity power spectrum $k^3P_{vv}(z=0)$ for $\mu=1, 0.5$ (solid black, cyan)  as predicted for General Relativity in the linear regime.
Overplotted are volume-weighted peculiar velocity shot-noise  $k^3\sigma^2/n$ at $z=0.1$ expected from 2- and 10-year (red, black) supernova densities, 0.08 and 0.15~mag (dashed, dotted) intrinsic magnitude dispersions, and TAIPAN (green, dash-dotted).
The bottom solid horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{max}$.
\label{power:fig}}
\end{figure}

@@

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields  absolute magnitude
dispersion $\sigma_M \gtrsim 0.12$~mag.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}. 
Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.  
At this precision the intrinsic velocity dispersion  at $z=0.028$ is  of $300$~km\,s$^{-1}$, i.e.\ a single SN~Ia  is of such quality as to
provide $S/N \sim 1$.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.  With similar densities,
 the (two) 10-year SN~Ia survey will have
 a (6) 29$\times$ reduction in shot-noise, $\sigma^2_M/n$, relative to the Fundamental Plane survey of TAIPAN.

@@

We project uncertainties  on the growth index, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}
\citep[there is an alternative approach using an estimator for the mean pairwise velocity][]{2011PhRvD..83d3004B}.
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_{r_{min}}^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the parameters considered are $\lambda \in \{\gamma,bD, \Omega_{M0}\}$.
The parameter dependence enters through $(fD)(\gamma)$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
Both $f$ and $D$ depend on $\Omega_M=\frac{\Omega_{M0}}{\Omega_{M0} + (1-\Omega_{M0})a^3}$.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered.

@@

We consider peculiar velocity surveys for a range of redshift depths $z_{max}$ for survey durations of $t=2$ and 10~years.
The other survey parameters
$\Omega=3\pi$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
The $k$-limits are  taken to be $k_{min} = \pi/r_{max}$  and 
$k_{max} = 0.1$~$h$\,Mpc$^{-1}$.
A minimum distance $r_{min}=r(z=0.01)$ is imposed as the physics derivation is taken at the limit
where peculiar velocities are significantly  smaller than
the cosmological redshift.
The sample-selection efficiency $\epsilon$ is redshift-independent, i.e.\ the native redshift distribution is not sculpted.
The input bias of SN~Ia host galaxies is set as $b=1.2$.  An independent measurement of $\Omega_{M0}=0.3\pm0.005$
is included and is important contributor to the $\gamma$ constraint.  Number densities are taken to be
direction-independent, neglecting the slight declination-dependence of SN-survey time windows.

@@

{\bf Host-galaxies Need Precise Redshifts:} Peculiar velocities come from the difference between host-galaxy and cosmological redshifts.  Fractional
host-redshift uncertainties
need to satisfy $\sigma_z/z \ll \sigma_M$ in order not to contribute significantly
to the error budget.   These redshifts could come naturally
from galaxy signal in  the spectroscopic transient classification provided the spectrograph had sufficient resolution $R \gg \sigma_M^{-1}$.  Alternatively, 
moderate-resolution (bright-)galaxy  redshift surveys (e.g.\ DESI, 4MOST) can  provide multiplexed observations of 
a significant fraction of nearby SN hosts before  discovery or after transient light has faded away.

@@

We have assumed that the galaxy and peculiar-velocity surveys as being the one and the same, since
the same galaxies are used for both overdensity and velocity measurements.  Though not considered in our calculations,
there would be benefit in increasing the overdensity sample to cover the larger volume that
affect the velocities within $z_{max}$, based on the expectations of linear theory and super-sample covariance.  Attaching these higher-redshift supernova hosts to
a peculiar-velocity survey would
not significantly strain the requirements, as they would not need precise distances.

@@

%SNe~Ia are already powerful probes of the homogeneous cosmological expansion of the Universe.  
In the next decade,
the high number of SN discoveries together with improved precision in their distance precisions will make $z<0.2$ SNe~Ia, more so than
galaxies,  powerful probes of gravity through their effect  on the growth of structure. 
A peculiar velocity survey must discover, classify, and get distances of SNe~Ia, as well as obtain redshifts of their host galaxies.
Choices in survey design impact the precision obtained on the growth index $\gamma$.

No other probe of growth of structure or tracer of peculiar velocity can alone provide comparable precision on  $\gamma$ in the next decade.
At low redshift, the RSD measurement is quickly sample variance limited (as are the planned DESI BGS and 4MOST surveys) making peculiar velocities the only 
precision probe of $fD$.
TAIPAN and a TAIPAN-like DESI BGS will be able to measure FP distances for nearly all usable nearby galaxies, so at low-$z$ the Fundamental Plane peculiar-velocity
technique will  saturate at a level that is not competitive with a modes 2-year SN survey.

Combined low-redshift peculiar velocity and high-redshift RSD $fD$ measurements are highly complimentary as together they probe the
$\gamma$-dependent shape of $fD(z)$ (not just its normalization) and potential scale-dependent influence of gravitational models, since low-
and high-redshift surveys are weighted by lower and higher $k$-modes respectively.
SN~Ia peculiar velocity surveys are of the highest scientific
interest and we encourage
the community to develop aggressive surveys in the pursuit of testing  General Relativity and probing gravity.

%\begin{comment}
\newpage
We are interested in
$$F^{-1}_{00,\alpha} = \frac{F_{11,\alpha}}{F_{00}F_{11}-F_{01}^2} - \frac{F_{11}}{(F_{00}F_{11}-F_{01}^2)^2}\left(F_{00,\alpha}F_{11}+
F_{00}F_{11,\alpha} - 2 F_{01}F_{01,\alpha}\right).$$
